---
title: "week-15: Indoor Air Pollution"
description: |
  Exploring share of deaths from indoor air pollution in Africa.
author:
  - name: Oumarou Diallo
date: 2022-04-14
preview: w15_indoor_air_pol.png
output:
  distill::distill_article:
    self_contained: false
    toc: true
---

## Setup & Libraries

```{r}
knitr::opts_chunk$set(echo = TRUE)

#if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidytuesdayR, tidyverse, janitor, scales, rmarkdown, ggrepel,
  ggthemes, MetBrewer, showtext, ggtext, patchwork, usefunc,
  sf, afrilearndata, sysfonts
  )

font_add_google("Noto Serif") 
font_add_google("Quicksand")
font_add_google("Fira Sans Condensed")
font_add_google('Roboto', 'Roboto')

showtext_auto()

```


## Datasets

This week, we have 6 datasets to play with. I explored them all but ended up focusing on 2: the fuel_gdp and indoor_pollution. 


Small issues with both datasets:
  - Some have very long names to work with (needed renaming)
  - For the fuel_gdp df, continent column has many missing (luckily easily fixable)

### Loading

```{r}
tt <- tt_load('2022-04-12')

```


### Exploring and wrangling 

```{r}
# Fuel_gdp 

fg <- tt$fuel_gdp %>% 
  clean_names() %>% 
  select(
    year,
    country = entity,
    pct     = access_to_clean_fuels_and_technologies_for_cooking_percent_of_population,
    gdp     = gdp_per_capita_ppp_constant_2017_international,
    pop     = population_historical_estimates,
    continent
    ) 
fuel_gdp <- fg %>% 
  mutate(continent2 = case_when(
    country %in% {fg %>% filter(continent == 'Asia') %>% 
        select(country) %>% pull()} ~ 'Asia',
    country %in% {fg %>% filter(continent == 'Europe') %>% 
        select(country) %>% pull()} ~ 'Europe',
    country %in% {fg %>% filter(continent == 'Africa') %>% 
        select(country) %>% pull()} ~ 'Africa',
    country %in% {fg %>% filter(continent == 'Antarctica') %>% 
        select(country) %>% pull()} ~ 'Antarctica',
    country %in% {fg %>% filter(continent == 'North America') %>% 
        select(country) %>% pull()} ~ 'North America',
    country %in% {fg %>% filter(continent == 'Oceania') %>% 
        select(country) %>% pull()} ~ 'Oceania',
    country %in% {fg %>% filter(continent == 'South America') %>% 
        select(country) %>% pull()} ~ 'South America')
    ) %>% 
  filter(
    !is.na(continent2),
    !is.na(pct)
    ) %>% 
  select(-continent) %>% 
  rename(continent = continent2)   # I hope to find a better/faster way to do this


# Indoor_pollution

ind_pollution <- tt$indoor_pollution %>% 
  clean_names() %>%
  rename(
    country = entity,
    deaths  = deaths_cause_all_causes_risk_household_air_pollution_from_solid_fuels_sex_both_age_age_standardized_percent)


```



### Final output


```{r}
paged_table(fuel_gdp)
```



```{r}
paged_table(ind_pollution)
```


## Plotting

### Correlation GDP - Access to Fuels + Tech


```{r}
p1 <- fuel_gdp %>% 
  filter(
    year      == 2015,
    continent == 'Africa', 
    !is.na(gdp)
    ) %>% 
  ggplot(aes(gdp, pct/100))+
  geom_point() + 
  geom_smooth(method = 'lm', fill = 'brown', alpha = .2)+
  geom_text_repel(
    aes(label = country), 
    segment.colour = NA)+
  scale_size_binned(range = c(1, 10))+
  scale_x_log10(labels = dollar)+
  scale_y_continuous(labels = percent) +
  theme_light()+
  labs( x        = 'GDP per capita',
        y        = 'Percentage',
        title    = 'Access to clean fuels & technologies for cooking in Africa',
        subtitle = 'GDP per capita vs. Population population with access in 2015')+
  theme(plot.title    = element_text(size = 20, face = 'bold', family = 'Noto Serif'),
        plot.subtitle = element_text(size = 12, face = 'bold', family = 'Quicksand'),
        axis.title    = element_text(size = 12, face = 'bold'),
        axis.text     = element_text(size = 10, face = 'bold'),
        panel.grid    = element_blank())


```



### Indoor air pollution

For this plot, I used the afrimapr data to plot the African continent with most up-to-date borders :)  To learn more =>  [afrimapr](https://afrimapr.github.io/afrimapr.website/blog/2020/healthsites-app/)



```{r}
data(africountries)
africa_data <- africountries %>% 
  dplyr::select(1, 6, 7, 8) %>% 
  rename(code = iso_a3) %>% 
  left_join(ind_pollution) %>% 
  dplyr::select(1,2,3,5,6,7,4) %>% 
  mutate(deaths = round(deaths, digits = 0))

# additional data prep for the legend
africa_data$deaths_cat <- cut(
  africa_data$deaths , 
  breaks         =c(0, 3, 6, 9, 12, 15, Inf), 
  labels         =c('0  -  3', '4  -  6', '7  -  9', '10-12', '13-15', '16-18'), 
  include.lowest = TRUE )

# countries with higher percentage in 2019
countries_list <- c('NER', 'BFA', 'TCD', 'ETH', 'MDG', 'COD', 'SOM', 'GIN')


# Plot title + subtitle 
# usefunc by [Nicola Rennie](https://twitter.com/nrennie35) is very handy for long text

mysubtitle <- str_wrap_break("Niger, Somalia, Chad, Congo D., Burkina Faso, Ethiopia & Guinea still have high share of deaths from indoor air pollution (>= 13%) in 2019\n\n", 70)
mycaption <- "W15 | Data: Our World in Data | #TidyTuesdayR"


# Another favorite package for theme+colors [metBrewer](https://github.com/BlakeRMills/MetBrewer)

p2 <- africa_data %>% 
  mutate(lab = case_when(code %in% countries_list ~ name,
                         TRUE ~ '')) %>% 
  filter(year %in% c(1990, 2000, 2010, 2019)) %>% 
  ggplot(aes(fill = deaths_cat)) +
  geom_sf(color = 'black', size = 0.05) +
  geom_sf_text(aes(label = lab), 
               color = 'white',
               size  = 3) +  
  facet_wrap(~year) +
  theme_map() +
  scale_fill_manual( 
    values=met.brewer("Tam", 8)[c(1, 2, 4, 5, 6, 7)], 
    name='', 
    guide = guide_legend(
      keywidth       =unit(6, units = 'mm'),
      keyheight      =unit(2, units = 'mm'),
      label.position = 'left', 
      title.position = 'top', 
      label.hjust    = -.01,
      ncol           =1)
  ) +
  plot_annotation(
    title    = 'INDOOR AIR POLLUTION',
    subtitle = mysubtitle,
    caption  = mycaption) &
  theme(
    text              = element_text(color = 'white'),
    legend.position   = c(0, 0),
    legend.background = element_rect(fill = NA, color = NA),
    legend.text       = element_text(face = 'bold', size = 12), 
    plot.margin       = margin(rep(4, 4)),
    strip.text        = element_text(face = 'bold', size = 20, color = 'white'),
    strip.background  = element_rect(fill = NA, color = NA),
    plot.background   = element_rect(fill  = met.brewer("Isfahan2")[5], color = NA), 
    panel.background  = element_rect(fill = met.brewer("Isfahan2")[5], color = NA),
    plot.title = element_text(family = "Fira Sans Condensed", face = "bold", hjust = 0.5, 
                              size = 24, color = "wheat", margin = margin(b = 10)),
    plot.subtitle = element_text(family = "Roboto", hjust = 0.5, 
                                 size = 12, color = "wheat"),
    plot.caption = element_text(hjust = .5, face = 'italic')
    )
  


```

